---
title: "Lab 6"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, error = TRUE,
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter the names of the group members here: EJ Porras, Michael Minton, YJ Chung

**This assignment is due by the end of the lab. Only one student in the group submits a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

In this lab, you will explore the `flights` dataset which comes with `nycflights13`. Let's first load the `tidyverse` package that contains `tidyr`, `dplyr` and `ggplot2` which we will need today:

```{r, warning=FALSE}
# If working on your own computer, remember to install new packages with install.packages("name") in the console
library(tidyverse)
library(nycflights13)
library(viridis)
```

Then let's take a quick look at the dataset:

```{r}
# Take a quick look
as_tibble(nycflights13::flights)
```

The nycflights13 dataset contains a bunch of dataframes about flights into and out of NYC in 2013.  We've used it in class, so you should be familiar with it!

The goal of the lab is to explore how dates work in R.

------------------------------------------------------------------------

### Question 1: (3 pts)

Start by extracting hour and month from the time_hour object.  (Flights already does this but that's not always the case, so it's a useful exercise.)

How does the distribution of flight times between 6 and 10 am change over the first 6 months of the year?  Since there are a bunch of lines, choose a color palatted where it's easy to see the difference between the months.  I recommend the viridis palatte. 

```{r}
# extract hours and month from time_hour and create plot of flight time dist'n
flights %>%
  mutate(Month = lubridate::month(time_hour), 
         Hour = lubridate::hour(time_hour)) %>%
  filter(Hour >= 6 & Hour <= 10) %>%
  filter(Month <= 6) %>%
  ggplot(aes(x = Hour)) +
  geom_bar(aes(fill = Month)) +
  facet_wrap(vars(month)) +
  scale_color_viridis(discrete = T)
```

**The distribution of flight time between 6 am and 10 am shifts slightly over the first six months of the year. As the month increases and one gets further into the year, there is a greater amount flights leaving around 6 am and less flights leaving around 8 am. The relative distribution of flights leaving at 7 am, 9 am, and 10 am remains the same throughout the first six months of the year.**


------------------------------------------------------------------------

### Question 2: (4 pts)

Find the most common tail number to fly into and out of NYC.  What date did it fly out of NYC the most?  What airline does it belong to (merge with airlines data to find out!)?  What is its most common destination? 

```{r}
# find most common tail number and its most common date, airline, and destination

# find the most common tail number to fly into and out of NYC
flights %>%
  filter(!is.na(tailnum)) %>%
  group_by(tailnum) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_max(count, n = 1)

# 
flights %>%
  filter(tailnum == "N725MQ") %>%
  mutate()


  group_by(date) %>%
  summarise(n = n()) %>%
  slice_max(date, n = 1)
```

**The most common tail number to fly in and out of NYC is N725MQ.**


------------------------------------------------------------------------

### Question 3: (4 pts)

Using the `weather` dataset and `time_hour` variable, plot the temperature at JFK for the months of January, May, September, and December.  Plot in different facets.  Edit the "scale" argument to facet_wrap such that the y axis is fixed and the x axis varies. Give the plot a title, axis labels, and remove the legend. 
(HINT: <https://ggplot2.tidyverse.org/reference/facet_wrap.html>)

```{r}
# your code goes below (make sure to edit comment)
weather
  #theme(legend.position="none")
```


### Question 4: (3 pts)

Now plot the data for the whole year by airport putting each airport in a different facet.  Use the facet_wrap option 'ncol = 1' to stack the facets vertically.  Use a color scale that makes cold temperatures blue and warm temperatures sienna2.  Also use "scale_x_datetime" to make the tick marks appear for every month and label them appropriately.  Are there any outliers that should be investigated?

```{r}
# your code goes below (make sure to edit comment)
weather

```


**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 5: (2 pts)

On what day of the week should you leave if you want to minimise the chance of a delay?  Does the answer differ if you want the shortest average delay?  HINT1: use the lubridate package to extract weekdays and make sure to set "label = T".  HINT2: Calculate percent delayed where the denominator is the total number of flights in a given weekday and the numerator is the number of flights that are delayed. HINT3: make delay a binary variable so all flights with delays get a 1 and all flights without get a 0.

```{r}
# your code goes below (make sure to edit comment)

flights 

```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 6: (4 pts)

Investigate an oddity - do any flights arrive before they depart?  If so, what's going on?

```{r}
# your code goes below (make sure to edit comment)
flights 
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 7: (4 pts) ###

We know that unique planes (i.e. tailnumbers) frequently fly into and out of NYC.  Pooling all of the airports together, what is the longest time between a plane taking off and landing in NYC and what is the shortest?  For the shortest, what is (likely) going on?

Hints: 
1) to do this, you need to arrange the data by `tailnum` and `time_hour`
2) you need to create a time difference using the lag() function.  lag() gives us the prior value in a dataframe/tibble
3) to display the difference not in seconds, use the function seconds_to_period from the lubridate package
4) to figure out what's going on, you need to filter by datetime.  To do that, make sure you add 'tz = "America/New_York")' as an argumnet

```{r}
# your code goes below (make sure to edit comment)

#get longest time

#get shortest time

#investigate shortest time

```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!
