---
title: "HW 2"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options=list(max.print=100))

# Edit the file starting below
```

## Enter your name and EID here: EJ Porras, ejp2488

**This homework is due on September 15, 2023 by 11:59pm. You will submit a pdf file on Gradescope.**

*For all questions, include the R code that you used to find your answer (show R chunk and outputs). Answers without supporting code will not receive credit (unless no code was required). Outputs without comments will not receive credit either: Write full sentences to describe your findings.*

------------------------------------------------------------------------

### Question 1: (2 pts)

##### The dataset `ChickWeight` contains information about the weights (in grams) of chicks on four different diets over time (measured at 2-day intervals) as the result of an experiment. The first few observations are listed below.

```{r}
# Save dataset as a dataframe
ChickWeight <- as.data.frame(ChickWeight)

# Visuaize the first ten rows of the dataset
head(ChickWeight, 10)
```

Use some combination of `table()` and `length()` to answer the following questions: 

- How many distinct chicks are there?  
- How many distinct time points?  
- How many distinct diet conditions?  
- How many chicks per diet condition?  

```{r}
## find nunber of chicks, time points, diet conditions, and chicks per diet condition
length(table(ChickWeight$Chick))
length(table(ChickWeight$Time))
length(table(ChickWeight$Diet))

length(unique(ChickWeight$Chick[ChickWeight$Diet == 1]))
length(unique(ChickWeight$Chick[ChickWeight$Diet == 2]))
length(unique(ChickWeight$Chick[ChickWeight$Diet == 3]))
length(unique(ChickWeight$Chick[ChickWeight$Diet == 4]))
```

**There are 50 distinct chicks separated into 4 distinct diet conditions and whose weight are measured at 12 distinct time points. There are 20 chicks on diet condition 1 and 10 chicks on diet conditions 2-4.**

------------------------------------------------------------------------

### Question 2: (2 pts)

##### Using the `ggplot2` package, create a simple scatterplot showing chick `weight` (on the y-axis) as a function of `Time`. Label the axes including the units of the variables and give the plot a title. How does chick `weight` change over `Time`?

```{r}
# Load the tidyverse (including ggplot2)
library(tidyverse)

## create scatterplot of weight as function of time
ChickWeight %>% 
  ggplot(aes(x = Time, y = weight)) + 
  geom_point(size = 3) +
  labs(title = "Relationship between Chick Weight and Days Since Birth",
       x = "Time (Days Since Birth)",
       y = "Weight (in grams)")
```

**Based on this scatterplot, we can see that there is a positive relationship between weight and time elapsed. Chicks gain more weight the longer they are alive. More interestingly, the variation in chick weight gets larger as time increases. While the spread of chick weights are very small for, we can see that the range of weights increases for larger values of time. For example, the range of chick weights at t = 4 is around 30 mg, while the range of chick weights at t = 20 is around 300 mg.**

------------------------------------------------------------------------

### Question 3: (2 pts)

##### Building upon the previous plot, map `Chick` to an aesthetic that assigns a color to each chick’s data points. Add lines that connect each chick’s points together with `geom_line()`. Finally, remove the legend. Do all chicks seem to gain weight in the same manner? Why/Why not?

```{r}
## scatterplot of weight as function of time with color for each chick and a line connecting each chick
ChickWeight %>% 
  ggplot(aes(x = Time, y = weight, color = Chick)) + 
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Relationship between Chick Weight and Days Since Birth",
       x = "Time (Days Since Birth)",
       y = "Weight (in grams)") +
  theme(legend.position = "none")
```

**It doesn't seem that all chicks gain weight in the same manner. There are slight variations in the weight gain patterns up until day 10, and the notion that they don't grow in the same manner becomes a lot more obvious part for all time observations past that point. Additionally, chicks don't grow weight at a steady clip; some plateau or even lose weight between observations.**

------------------------------------------------------------------------

### Question 4: (2 pts)

Continue modifying the same plot by 

* removing the color from the points only
* make all of the points white 
* leave the lines colored by chick
* Put the points *on top of* the lines 

On which day was the last value of the chicks' weight recorded?

```{r}
## scatterplot from Q3 with white points on top of the colored lines
ChickWeight %>% 
  ggplot(aes(x = Time, y = weight, color = Chick)) + 
  geom_line() +
  geom_point(color = "white", size = 1) +
  labs(title = "Relationship between Chick Weight and Days Since Birth",
       x = "Time (Days Since Birth)",
       y = "Weight (in grams)") +
  theme(legend.position = "none")
```

**After changing the color of the points to white and overlaying on top of the lines colored by chick, the last value of the chicks' weight were recorded on the 21st day, which is three weeks after they were born.**

------------------------------------------------------------------------

### Question 5: (2 pts)

Now, facet this plot by diet. Can you tell from this new plot which diet results in greater weight? Explain.

```{r}
## create scatterplot from Q4 faceted by diet
ChickWeight %>% 
  ggplot(aes(x = Time, y = weight, color = Chick)) + 
  geom_line() +
  geom_point(color = "white", size = 1) +
  facet_wrap(vars(Diet))
  labs(title = "Relationship between Chick Weight and Days Since Birth",
       x = "Time (Days Since Birth)",
       y = "Weight (in grams)") +
  theme(legend.position = "none")
```

**After faceting the scatterplot we created in question 4 by diet, we can see that diet #3 results in the greatest weight. The highest weight gains come from chicks who ate from the third diet. The average growth for chicks under the 3rd diet is similar to the highest weight gain from other diets. Even the smallest weight gains in diet #3 is comparable to the average weight gain experienced by chicks in other diets.**

------------------------------------------------------------------------

### Question 6: (2 pts)

Go back to your plot from question 2 and fit a *linear regression line* (using `lm`) to the chicks in each diet with `geom_smooth()`. There should be 4 separate regression lines, one for each diet, each a separate color. Can you see more clearly which diet results in greater weight? Explain.

```{r}
## create the scatterplot from Q2 with linear regression lines for each diet
ChickWeight %>% 
  ggplot(aes(x = Time, y = weight, color = Diet)) + 
  geom_smooth(method = "lm") +
  geom_point(size = 1) +
  labs(title = "Relationship between Chick Weight and Days Since Birth",
       x = "Time (Days Since Birth)",
       y = "Weight (in grams)")
```

**Whereas we merely assumed that diet #3 resulted in greater weight among the chicks in question 5, adding these 4 regression lines by diet clearly shows that diet #3 resulted in the greatest weight gain among newborn chicks. When we look at the estimated weight of a typical chicks for each of the four diets at day 21, we see that the highest estimated weight comes from diet #3. Additionally, diet #3 has the greatest slope in magnitude.**

------------------------------------------------------------------------

### Question 7: (2 pts)

A scatterplot might not be the best way to visualize this data: it calls attention to the relationship between weight and time, but it can be hard to see the differences between diets. A more traditional approach for exploring the effect of diet would be to construct a barplot representing group means with standard error bars showing +/- 1 standard error.

Create a plot using `geom_bar()` where each bar's height corresponds to the average chick weight for each of the four diet conditions. Rename the y-axis to include units (e.g., with scale_y_continuous(`name=`...)) and make the major tick marks go from 0 to 150 by 25 (e.g., with scale_y_continuous(`breaks=`...)). Which diet has the highest mean `weight`? 

```{r}
## create barplot with bar height as average chick weight per diet condition
ChickWeight %>%
  ggplot(aes(x = Diet, y = weight)) + 
  geom_bar(stat = "summary", position = "dodge", fun = mean) + 
  scale_y_continuous(name = "Weight (in grams)", breaks = seq(0, 150, by = 25))
```

**After creating this bar graph, we can see that diet #3 has the highest mean 'weight'.**

------------------------------------------------------------------------

### Question 8: (2 pts)

Add error bars showing + or - 1 SE using `geom_errorbar(stat = "summary")`. Make the error-bars skinnier by adding a `width =`**0.5** argument. Color the bars (not the error bars, but the barplot bars) by diet and change from the default color scheme using a `scale_fill_` or a `scale_color_`. diet seems to have the most variation in `weight`? The least variation?

```{r}
## create bar graph from Q7 with error bars of width 0.5 and color scale by diet
ChickWeight %>%
  ggplot(aes(x = Diet, y = weight)) + 
  geom_bar(aes(fill = Diet), stat = "summary", position = "dodge", fun = mean) +
  geom_errorbar(stat = "summary", width = 0.5) +
  scale_fill_brewer(palette = "Oranges") +
  scale_y_continuous(name = "Weight (in grams)", breaks = seq(0, 150, by = 25))
```

**Diet #3 seems to have the most variation because it has the largest range for its error bars, while diet #1 has the least variation because it has the smallest range for the error bars.**

------------------------------------------------------------------------

### Question 9: (2 pts)

Take your code from question 8 and replace `geom_bar()` with `geom_point()`. Remove the `breaks=` argument from `scale_y_continuous`. Make the points larger and color them all red. Put them *on top of* the error bars. Does the mean chick weight seem to differ based on the diet? *I am not asking to conduct hypothesis testing but informally state if they seem to differ and if so, how.*

```{r}
## create scatterplot using the bar graph from Q8 as a reference
ChickWeight %>%
  ggplot(aes(x = Diet, y = weight)) + 
  geom_errorbar(stat = "summary", width = 0.5) +
  geom_point(stat = "summary", position = "dodge", fun = mean, size = 4, aes(color = "red")) + 
  scale_y_continuous(name = "Weight (in grams)") +
  theme(legend.position = "none")
```

**This scatterplot suggests that the mean chick weight seems to differ depending on the diet that the chick receives. In particular, it seems that diet #3 is the most effective since it has a higher concentration of chicks with higher weights while also having the largest variation, suggesting that there is a big concentration of chicks that weigh more than their counterparts.**

------------------------------------------------------------------------

### Formatting: (2 pts)

##### Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------