---
title: "HW 6"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter your name and EID here: EJ Porras, ejp2488

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

We will use the packages `tidyverse`, `factoextra`, and `cluster` for this assignment.

```{r, message = FALSE}
# Load packages
library(tidyverse)
library(factoextra)
library(cluster)
```

------------------------------------------------------------------------

## Question 1: (2 pts)

The dataset for this homework comes from the article:

*Tsuzuku N, Kohno N. 2020. The oldest record of the Steller sea lion Eumetopias jubatus (Schreber, 1776) from the early Pleistocene of the North Pacific. <https://doi.org/10.7717/peerj.9709>*

Read the ***Abstract*** of the article and the section called *Results of Morphometric Analyses*. What was the goal of this study and what was the main finding?

**The goal of this study is to determine the evolutionary history of GKZ-N 00001 since it is the earliest fossil that has been identified as a species of the sea lion genus *Eumetopias*. After comparing morphometric comparisons among 51 mandibles of fur seals and sea lions with GKZ-N 00001, researchers find that there little variation between GKZ-N 00001 and male *E. jubatus* in terms of mandible measurements and canine root proportions. Overall, this seems to suggest that GKZ-N 00001 can be identified as the Stellar sea lion *E. jubatus.* **

------------------------------------------------------------------------

## Question 2: (1 pt)

Under the supplemental information, I retrieved the data from a word document into a `.csv` document. Import the dataset from GitHub. 

```{r}
# download data from GitHub
sealions <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//Sealions.csv")
```

How many rows and how many columns are in this dataset? What does a row represent? What does a column represent?

```{r}
# find number of rows and columns in dataset
nrow(sealions)
ncol(sealions)
```

**There are 51 rows and 39 columns in the dataset. A typical row in this dataset represents a fossil (specifically a mandible) for a given gender and species that may be GKZ-N 00001, while a typical column represents a specific measurement that the researchers took for the mandible of these fossils.**

------------------------------------------------------------------------

## Question 3: (1 pt)

Before we can analyze the data, let's do some cleaning. Using a combination of the `select()`, `where()`, and a predicate function like `is.character()` we can scan through all the columns of the dataset and see which columns are of character type.

```{r}
## Select all of the columns that have character type
sealions %>% 
    select(where(is.character))
```

When importing this dataset into RStudio, which variables were considered numeric? Why are some measurements not considered as numeric? Use the `is.numeric()` predicate function here.


```{r}
# see what variables are considered numeric in dataset
sealions %>%
  select(where(is.numeric))
```

**The variables "K. and "AD" are considered numeric in this dataset. When looking at the dataset, these are the only variables that have a value for every entry. The other variables have at least one entry marked as "-". This may explain why some variables are considered numeric, and other variobles are considered as characters.**

------------------------------------------------------------------------

## Question 4: (1 pt)

The functions `mutate()` and `across()`, when used together, can make changes across a range of columns in a data frame. 

Using `mutate()` and `across()`, replace all `-` in the dataset with the missing values *NA* and then then make sure all measurements are defined as numeric variables with. The first part of the code replaces the "-" with NA values. Write the second part to coerce all of the columns (except for the ID column!) to be numeric. Make sure to overwrite the dataset `sealions`. 

NOTE: Look at the examples in `?across` to get a sense of how to use the `across()` function.

```{r}
# replace "-" to NA and change measurement columns to numeric variables
sealions <- sealions %>% 
    ## Replace all "-" with NA
    mutate(across(where(is.character), ~ na_if(.x, "-"))) %>% 
    ## Coerce all columns (except for ID) to be numeric
    mutate(across(where(is.character) & !c(ID), as.numeric))
```

What is the mean rostral tip of mandible `C`?

```{r}
# find mean rostral tip of mandible 'C'
sealions %>%
  summarise(mean_rostral_tip = mean(C, na.rm = TRUE))
```

**The mean rostral tip of mandible C is 34.86622 mm.**

------------------------------------------------------------------------

## Question 5: (2 pts)

You are given the code in this question. But what does the code do? Write comments.

```{r}
sealions <- sealions %>% 
  ## removes columns where there is no recorded measurement for species GKZ-N 00001
  select_if(!is.na(sealions[51,])) %>% 
  ## removes all observations where there is at least one NA value for a measurement
  na.omit
```

How many columns and how many rows are remaining in this dataset?

**After running this block of code, there are 42 rows and 23 columns remaining in the dataset 'sealions.'**

------------------------------------------------------------------------

## Question 6: (2 pts)

Use `dplyr` functions on `sealions` to split the `ID` variable into two variables `species` and `sex` with the function `separate()`. *Hint: in the `ID` variable, what symbol separates the species from sex?* The article states that the fossil specimen has to be male. Replace the missing value of `sex` for the fossil specimen GKZ-N 00001. *Hint: You could use the functions `mutate()` and `replace_na()`.* Save the resulting dataset as `sealions_clean`.

```{r}
# separate ID into species and sex and include sex of fossil specimen GKZ-N 00001
sealions_clean <- sealions %>%
  separate(col = "ID", c("species", "sex"), sep = "\\[|\\]") %>%
  mutate(sex = replace_na(sex, "m"))
```

How many sealions are male/female?

```{r}
# find number of male and female sealions
sealions_clean %>%
  count(sex)
```

**In the cleaned sealions dataset, there are 23 female specimens and 19 male specimens.**

------------------------------------------------------------------------

## Question 7: (1 pt)

Using `dplyr` functions, only keep numeric variables and scale each numeric variable. Save the resulting dataset as `sealions_num`. What should the mean of the scaled variable of the rostral tip of mandible C be?

```{r}
# keep numeric variables and scale each numeric variable
sealions_num <- sealions_clean %>%
  select(-is.character) %>%
  scale()
```

**The mean of the scaled variable of the rostral tip of mandible C should be 0 since we are normalizing all rostral tip measurements of mandible C relative to the mean.**

------------------------------------------------------------------------

## Question 8: (2 pts)

Let's  perform PCA on the measurements available for the fossil specimen GKZ-N 00001. Using the function `prcomp()`, calculate the principal components (PCs) for the scaled data, `sealions_num`, obtained in the previous question. Construct a scree plot with the function `fviz_eig()` from the package `factoextra`. What is the cumulative percentage of explained variance for PC1 and PC2? 

```{r}
# construct scree plot for principal components of sealions_num
fviz_eig(prcomp(sealions_num))
summary(prcomp(sealions_num))
```

**After creating a scree plot and reporting the summary statistics for PCA, the cumulative percentage of explained variance for PC1 (0.8476) and PC2 (0.04704) is 89.464%.**

------------------------------------------------------------------------

## Question 9: (2 pts)

How many *known species* are there in `sealions_clean`? Therefore, how many clusters should we look for to identify what species GKZ-N 00001 most likely belongs to? 

```{r}
# find number of known species in 'sealions_clean'
sealions_clean %>%
  filter(species != "GKZ-N 00001") %>%
  group_by(species) %>%
  summarise(distinct_species = n_distinct(species))
```

**There are 3 known species in sealions_clean. Thus, we are looking for 3 clusters (1 for each species) in order to identify what species GKZ-N 00001 most likely belongs to.**

Try using the PAM clustering algorithm using the `pam()` function in R (it is similar to K-means). Perform the PAM clustering algorithm on `sealions_num`, run the PAM clustering algorithm.

```{r}
# perform PAM clustering algorithm on 'sealions_num'
pam_results <- pam(sealions_num, 3) 
```

------------------------------------------------------------------------

## Question 10: (2 pts)

Represent the clusters along the first two principal components and specify to shape the observations by their `species` in the aesthetics. *Note: you can either use `ggplot` or `fviz_cluster()`.* 

```{r}
# create cluster visualization of first two principal components
fviz_cluster(pam_results, sealions_num, geom = NULL) +
  geom_point(aes(shape = sealions_clean$species)) +
  labs(title = "PAM Clustering by Species", 
       shape = "Sealion species")
```

The fossil specimen GKZ-N 00001 appears to be close to which species? 

**The fossil specimen GKZ-N 00001 appears to be close to the species E. jubatus, since the GKZ-N 00001 makes up one of the boundaries for cluster 1, which is composed of observations of E. jubatus.**

------------------------------------------------------------------------

## Question 11: (2 pts)

Putting it all together. Reflect on and summarize in 1-2 sentences the different steps taken through this assignment. Compare your conclusions to the findings discussed by the researchers in the article (cite their findings).

**Given the sealions dataset, we cleaned it by changing measurement data to numeric variables, removing NA columns and observations, splitting recorded observations by species and sex, and scaling the data to normalize measurements. We then conducted clustering analysis to determine what species GKZ-N 00001 is most likely related to, finding that it is most like the species E. jubatus. This mirrors the findings found by the researchers, who identified GKZ-N 00001 as E. jubatus based on the morphometric analyses, which consisted of bivariate analysis, an inquiry into the proportion of the canine root, and PCA analysis.**

------------------------------------------------------------------------

## Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```
