---
title: "Lab 4"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter the names of the group members here: EJ Porras, Michael Minton, YJ Chung

**This assignment is due by the end of the day Only one student in the group submits a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

In this lab, you will explore the dataset `starwars` which comes with `dplyr`. Let's first load the `tidyverse` package that contains both `dplyr` and `ggplot2` which we will need today:

```{r, warning=FALSE}
# If working on your own computer, remember to install new packages with install.packages("name") in the console
library(tidyverse)
```

Then let's save the dataset in our environment and take a quick look at it:

```{r}
# Upload data from GitHub
starwars <- as_tibble(starwars)

# Take a quick look with glimpse() 
glimpse(starwars)

# or learn more by running ?starwars in the console
```

This dataset contains information about Starwars characters: name, height, mass, hair_color, skin_color, eye_color, birth_year, sex, gender, homeworld, species, films, vehicles, starships.

The goal of the lab is to investigate some characteristics of the Starwars characters using `dplyr`.

------------------------------------------------------------------------

### Question 1: (3 pts)

Using `dplyr` core functions, create a new variable in the data frame to calculate Body Mass Index (BMI). BMI is for a height in **meters** and a mass in **kilograms**: 

> BMI = mass / height^2.

Only display the top 5 observations for BMI and just keep relevant information (name, species, height, mass). Who has the highest BMI in the dataset?

```{r}
# create BMI variable and display top 5 observations for BMI
starwars %>%
  mutate(name, BMI = mass / (height/100) ** 2) %>%
  select(name:mass, species, BMI) %>%
  slice_max(BMI, n = 5)

```

**Jabba Desilijic Tiure, a Hutt, has the highest BMI in the dataset with a BMI of 443.**

------------------------------------------------------------------------

### Question 2: (3 pts)

Using `dplyr` core functions, find how many characters' names there are per species. Which species contain the two largest numbers of character names?

```{r}
# find number of character names per species
starwars %>%
  group_by(species) %>%
  count(species)
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 3: (3 pts)

Using `dplyr` core functions and `ggplot`, compare the distributions of BMI between the two species found in the previous question using boxplots and simultaneously showing the data with `geom_point()`. Which of the two species seem to have the highest BMI on average? Is that reasonable to make such a comparison?

```{r}
# your code goes below (make sure to edit comment)

```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 4: (3 pts)

Let's look at who has no hair across planets! 

First, remove the NAs from hair_color. *Hint: use `!` and `is.na()` in a `dplyr` core function.*

Second, find the mean of a logical vector inside `summarize()` to get the proportion of bald characters in **each** homeworld. Call the resulting proportion `pBald`. *Hint: `hair_color == "none"` designates bald characters.*

How many home worlds only host bald characters?

```{r}
# your code goes below (make sure to edit comment)

```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 5: (4 pts)

Choose one of the research questions proposed by your group members in the Pre-Lab Quiz. Answer your research questions using at least 2 of the 6 `dplyr` core functions and with a `ggplot` visualization. Include a headline to your visualization and interpret what you see!

```{r}
# your code goes below (make sure to edit comment)

```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 6: (2 pts)

What did you expect the answer to this research question to be before you looked at the data? How does the result compare to what you expected?

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!
