---
title: "Lab 6"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, error = TRUE,
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter the names of the group members here: EJ Porras, Michael Minton, YJ Chung

**This assignment is due by the end of the lab. Only one student in the group submits a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

In this lab, you will explore the `who` dataset which comes with `tidyr`. Let's first load the `tidyverse` package that contains `tidyr`, `dplyr` and `ggplot2` which we will need today:

```{r, warning=FALSE}
# If working on your own computer, remember to install new packages with install.packages("name") in the console
library(tidyverse)
```

Then let's take a quick look at the dataset:

```{r}
# Take a quick look
as_tibble(who)
```

The `who` dataset contains information about tuberculosis (TB) cases per country over the years. The TB cases are reported in the columns `new_sp_m014:newrel_f65` following these conventions:

1.  The first three letters of each column denote `new` cases.

2.  The next two/three letters describe the type of TB: `rel` = cases of relapse, `ep` = cases of extrapulmonary TB, `sn` = cases of pulmonary TB that could not be diagnosed by a pulmonary smear, `sp` = cases of pulmonary TB that could be diagnosed be a pulmonary smear.

3.  The next letter indicates the gender category: females `f` or males `m`.

4.  The remaining numbers gives the age group (for example, 014 means 0-14 years old).

The goal of the lab is to compare tuberculosis (TB) cases across countries over time, first tidying the data.

------------------------------------------------------------------------

### Question 1: (3 pts)

Consider the `who` dataset. Using  `tidvyerse` functions, convert the dataset into a long format by collecting all of the `new_diagnosis_genderage`-columns into a single column (call it `new_diagnosis_genderage`) and all of their values into another single column (call it `cases`). Save the resulting dataset as an R object called `long_who`. How many rows does the `long_who` dataset have?

```{r}
# convert dataset into long format and save as new long_who
long_who <- who %>%
            pivot_longer(cols = new_sp_m014:newrel_f65, 
                         names_to = "new_diagnosis_genderage", 
                         values_to = "cases")

nrow(long_who)
```

**After converting the dataset into a long format by collecting all of the 'new_diagnosis_genderage' columns into a single column and all of their values into another single column, we can see that the 'long_who' dataset has 405,440 rows.**

------------------------------------------------------------------------

### Question 2: (4 pts)

Next, we would like to separate the `new_diagnosis_genderage` variable into `diagnosis`, `gender`, and `age`. Let's use the `separate()` function. However, there is a discrepancy in coding for new cases of relapse: they appear as `newrel` while all other types of cases use an `_` after `new`. The `mutate` function below will take care of that.

Apply `separate()` to split the variable `new_diagnosis_genderage` into 3 variables called `new`, `diagnosis`, and `genderage`. Then apply `separate()` again to split `genderage` into 2 variables called `gender` and `age`. You will need to specify the separator this time as there is no special character! *Hint: use `sep =` a certain number which corresponds to the number of characters after which to make the split.* Finally, drop the variable `new` since we only have new cases. Call the resulting dataset `tidy_who`. How many columns are left in `tidy_who`?

```{r}
tidy_who <- 
    long_who %>%
    mutate(new_diagnosis_genderage = 
               str_replace(new_diagnosis_genderage, "newrel", "new_rel")) %>%
  # separate new_diagnosis_genderage into new, diagnosis, gender, and age
    separate(col = "new_diagnosis_genderage", c("new", "diagnosis", "genderage")) %>%
    separate(col = "genderage", c("gender", "age"), sep = 1) %>%
    select(-"new")
    
ncol(tidy_who)
```

**After separating 'new_diagnosis_genderage' into new, diagnosis, gender, and age and dropping 'new', there are 8 columns left in 'tidy_who'.**

------------------------------------------------------------------------

### Question 3: (2 pts)

Let's take a look at missing values. The main concern is whether a missing value means that there were no cases of TB or whether it means that the WHO does not have data on the number of TB cases. Are there any zeros in our `tidy_who` dataset? What does it mean about `NA` values?

```{r}
# check to see if there are 0 in 'tidy_who'
tidy_who %>%
  filter(cases == 0) %>%
  nrow()

```

**Since there are 11,080 0's in the tidy_who dataset, we can assume that NA values mean that the WHO does not have enough information to say that a given country in a given year contains any evidence of TB cases in that area.**

------------------------------------------------------------------------

### Question 4: (3 pts)

What about missing years for some countries? Those would not appear explicitly in the dataset, they just would not be there. Using `dplyr` functions, find the total number of distinct years for each country in `tidy_who`. 

Also report the minimum and maximum year contained in the dataset for each country. How many countries had less than the expected 34 years (1980 to 2013)? Why do you think these years are missing? *Note: To understand why we have missing years, look at Serbia & Montenegro. What happened to this country?*

```{r}
# find number of distinct years, minimum year, and maximum year by country
tidy_who %>%
  group_by(country) %>%
  summarize(num_years = n_distinct(year),
            min_year = min(year), 
            max_year = max(year))
  
  
```

**Some years are missing for some countries because of geopolitical circumstances, such as civil war leading to some countries dissolving and/or unifying. An example of this would be Serbia & Montenegro, which became two separate countries in 2006.**

------------------------------------------------------------------------

### Question 5: (2 pts)

Investigate how the total number of TB cases (adding cases across all diagnoses) changes over the years splitting by gender or age categories. Write a research question/hypothesis to explore this change.

**Our research question is the following: how does the total number of TB cases vary by age category?**

------------------------------------------------------------------------

### Question 6: (4 pts)

To answer your research question, first find the annual total number of TB cases in `tidy_who` per age or gender category. Then use a `ggplot` visualization. Include a headline to your visualization and interpret what you see! *Note: remember to handle missing values!*

```{r}
# make barplot of number of TB cases by age category
tidy_who %>%
  group_by(age) %>%
  summarize(num_cases = sum(cases, na.rm = T)) %>%
  ggplot(aes(x = age, y = num_cases)) +
  geom_bar(stat = "identity") +
  labs(x = "Age Category", 
       y = "Number of Cases", 
       title = "Number of Tuberculoses Cases by Age Category")
```

**After making a barplot of the number of TB cases by age category, we can see that a lot of adults in the presumably healthy adult age ranges are more likely to contract TB. We can see that adults aged 25-34 have the most number of TB cases with the 35-44 and 15-24 age categories not too far behind. This surprises me somewhat given what we know about immunology because young children and the elderly are more likely to contract bad strains of most viral illnesses. I guess that the sheer number of adults in the population outweigh this fact, though.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!
